Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.

### 2023-11-08 scripts/manuscript/manuscript.qmd
final comments

---
title: Disentangling linkages between satellite-derived forest structure and productivity essential biodiversity variables.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices (DHIs)
  - essential biodiversity variables
  - redundancy analysis (RDA)
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: remote-sensing-of-environment.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Nature Scientific Reports
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
number-sections: true
---

@@

Key to regular biodiversity monitoring is the definition and selection
of variables capable of accurately monitoring biodiversity trends. One
framework is the Essential Biodiversity Variables (EBV), which are
standardized, comparable data that represents a minimum set of
biological information required for monitoring biodiversity change at
large spatial extents, with each class being complementary to one
another -- i.e., capturing unique biological information. The
interrelationships between forest ecosystem structure and function and
the complementarity between these two classes of EBVs across a range of
ecosystems encompassing significant environmental gradients are examined
using redundancy analysis (RDA) and variation partitioning. We found
that the primary and modelled forest structure variables are not
strongly related to the Dynamic Habitat Indices (DHIs), with the
structural information only explaining 16% of the variation in the DHIs,
indicating that structure and function EBVs are suitable to be used
together for monitoring forest ecosystems. We also found that forest
types and ecosystems dominated by coniferous trees had less variation
explained, potentially due to the temporal decoupling of canopy cover
and the DHIs. We suggest that biodiversity researchers focus on using
forest structural attributes derived directly from lidar data alongside
the DHIs as complementary EBVs for monitoring forest ecological
integrity.

@@

Monitoring the changing state of biodiversity globally is key to
understanding and mitigating the accelerated extinction risk for many
species [@thomas2004; @urban2015] and the homogenization of biotic
communities at various spatial and temporal scales
[@mcgillFifteenFormsBiodiversity2015]. The Group on Earth Observations
Biodiversity Observation Network has developed the Essential
Biodiversity Variables [EBVs, @pereira2013], which are standardized,
comparable data that represents a minimum set of biological information
required for monitoring biodiversity change at large spatial extents
[@skidmore2021]. The EBVs are designed as an analog to the Essential
Climate Variables framework [@bojinski2014] to be global in scope,
relevant to biodiversity information, feasible to implement and use, and
complementary to one another [@skidmore2021]. While it can be difficult,
time consuming, and expensive to collect data on biodiversity across
wide swaths of land, EBVs can be combined with sampled biodiversity
information to allow for the spatial extrapolation of more sparsely
collected in situ data [@cavender-bares2022]. The standardization and
global availability of the EBVs allows for the assessment and monitoring
of protected area effectiveness and ecosystem health at large spatial
scales [@hansen2021], which can then be integrated into policy
development [@geijzendorffer2016]. There are six EBV classes, each of
which provide information on a different facet of biodiversity: species
populations, species traits, community composition, ecosystem structure,
ecosystem function, and genetic composition [@pereira2013]. While these
EBV classes can be related to one another under a variety of ecological
and evolutionary mechanisms, they provide distinct information; multiple
in concert can provide a more complete picture on the status and trends
of biodiversity in a region.

The global coverage and moderate spatial resolutions of many optical
satellite remote sensing missions [@wulder2022], combined with open-data
policies [@wulder2012], make these datasets exceptionally well suited
for contributing to some EBV classes [@skidmore2021]. The six EBV
classes can be divided into two general approaches: species-focused and
ecosystem-focused [@fernández2020]. Genetic composition, species
populations, and species traits are species-focused EBVs, with community
composition being directly related to species populations. While
species-focused and community composition EBVs can be assessed from
measurements at the tree-crown scale using very-high spatial resolution
imagery [@fassnacht2016; @graves2016; @wu2021], the spatially
limited--and often *ad-hoc*--data collection associated with monitoring
individuals is not conducive to the global or regional scales often
required for biodiversity trend assessments [@valdez2023]. In contrast,
two ecosystem-based EBV classes (ecosystem structure and ecosystem
function) are well suited to be monitored across large extents using
moderate spatial resolution optical imagery, such as the Moderate
Resolution Imaging Spectroradiometer [MODIS, @zhang2003], Landsat
imaging systems [@wulder2022] or Sentinel-2 [@helfenstein2022;
@drusch2012] programs, especially in forested environments
[@pettorelli2018; @coops2016].

Ecosystem structure is the composition, abundance, and spatial
arrangement of different ecosystem components, and is indicative of
habitat quality and connectivity [@noss1990; @valbuena2020]. There are
both horizontal and vertical components of ecosystem structure. In
forest environments, forest structural diversity has been linked to
species richness at various spatial scales, ranging from individual
plots to landscapes [@bergen2009; @gao2014]. Structural attributes range
in complexity from simple (e.g., canopy cover, canopy height), to more
complex (e.g., foliage height diversity, leaf area index) to modelled
(e.g., aboveground biomass, basal area), which can be derived using
lidar (light detection and ranging) data [@coops2021]. A suite of these,
and other, structural attributes has been used as indicators of
biodiversity at local to landscape scales [@lefsky1999; @guo2017;
@coops2016; @bergen2009]. Increased forest structural complexity has
been hypothesized to create additional niches, leading to increased
species diversity [@bergen2009], which has been frequently demonstrated
using avian species diversity metrics [@macarthur1961], with lidar
[@clawges2008], a combination of lidar and satellite imagery
[@herniman2020], with some research indicating that canopy vertical
distribution was the strongest predictor of species richness
[@goetz2007]. Although lidar datasets have traditionally been local in
extent, advances in satellite remote sensing processing have allowed the
extension of three-dimensional forest structure attributes nationally in
a wall-to-wall fashion. The wall-to-wall data often uses data fusion
approaches involving lidar data and medium spatial resolution optical
imagery [@matasci2018; @matasci2018b; @coops2021].

Ecosystem functions are measures of ecosystem performance that are the
consequence of one or multiple ecosystem processes [@pettorelli2018;
@ecosyste2005]. Within the EBV framework, vegetation phenology and
metrics of primary productivity are commonly used as potential EBVs
[@skidmore2021]. Energy availability is the amount of accessible and
utilizable energy available for growth in an ecosystem, and is often
measured as net primary productivity [@bonn2004]. Energy availability
has shown to be a predictor of species richness and abundances at
various scales [@chase2002; @radeloff2019; @coops2019; @razenkova2023],
and is measurable using satellite remote sensing via various vegetation
indices [@huete2002; @radeloff2019]. Vegetation indices are indicative
of photosynthetic activity, and are commonly used as proxies of gross
primary productivity [@huang2019]. These vegetation indices enable the
assessment of biodiversity patterns at single time points [@bonn2004],
or through time when using time series data [@berry2007; @radeloff2019].

The relationship between energy availability and biodiversity can be
explained through various hypothesized mechanisms: including available
energy, environmental stress, and environmental stability. These three
hypotheses have been incorporated into the Dynamic Habitat Indices
(DHIs), which summarizes intra-annual dynamics in remotely-sensed
productivity metrics and have been evaluated as broad-scale biodiversity
indicators [@berry2007; @coops2008; @radeloff2019]. According to the
available energy hypothesis [@wright1983] an increase in available
energy results in higher species richness. In the context of the DHIs,
the total energy available within a specific area is quantified using
the cumulative DHI, which sums each productivity observation throughout
the course of a year. The environmental stress hypothesis [@currie2004]
suggests that higher levels of minimum available energy contribute to
greater species richness. The minimum DHI, derived from the lowest
productivity observed over the course of a year, indicates the level of
available energy during stressful periods. The environmental stability
hypothesis [@williams2008] proposes that lower energy variation
throughout the year leads to increased species richness. The variation
DHI captures the coefficient of variation in a vegetation index over the
course of a year, reflecting the stability of energy availability. The
DHIs have previously been produced at a global extent using MODIS
imagery and have been used to assess alpha [@radeloff2019] and beta
[@andrew2012] diversity, species abundances [@razenkova2023], and
construct novel ecoregionalizations [@coops2009; @andrew2013]. Recent
studies have increased the spatial resolution of these indices by
calculating the DHIs from multi-annual Landsat imagery
[@razenkovaMediumresolutionDynamicHabitatInPress].

Choosing datasets that are most closely related to the phenomenon of
interest in a given application allows for direct integration of
monitoring results with management actions [@pressey2021]. With the
advent of large-extent monitoring methods like satellite remote sensing
which have led to a proliferation of potential EBV datasets, it becomes
important to assess the interrelationships between these datasets and
the complementary of the information they provide to reduce redundancy
in future EBV development efforts [@pereira2013; @skidmore2021].
Identifying and exploring strong relationships between EBVs may help
elucidate the ecological relationships between these facets of
biodiversity, such as ecosystem structure and function, at local to
global scales. On the other hand, when variables are not related, they
may be well suited to be used in monitoring programs together, as
complementary EBVs.

Linkages between forest ecosystem structure and function have been
examined within a remote sensing context for decades [@huete2002,
@knyazikhin1998; @myneni1994]. While there is significant theoretical
and empirical evidence for their relationship [@myneni1994], the
direction and significance has not been consistent, with positive,
negative, and non-significant relationships being found [@ali2019].
Hypothesized mechanisms such as niche complementarity suggest that
productivity increases with stand structural complexity [@zhang2012],
while asymmetric competition for light can reduce forest productivity
with increased structural complexity [@bourdier2016]. Some aspects of
forest structure, such as canopy cover and leaf area index, also
directly influence vegetation indices, which are used to estimate
productivity, especially in very dense or very sparse forests
[@pettorelli2005]. Further, the relationship between forest structural
diversity metrics--which are now more accurately and comprehensively
derived from lidar data--and the temporal variation in functional
metrics, specifically the metrics of ecosystem productivity via the DHIs
framework, have yet to be fully examined.

The goal of this study is to examine relationships between forest
ecosystem structure and function, and the complementarity between these
two classes of EBVs across significant environmental gradients. We
synthesize data from moderate-scale remote sensing-derived metrics of
forest structure (simple lidar-extracted metrics of canopy height,
canopy cover, and structural complexity), as well as modelled forest
structure attributes (gross stem volume, basal area, and aboveground
biomass), with a well established remote sensing-derived index for
ecosystem function (DHIs). First, we analyze remote sensing metric
complementarity by examining the multivariate relationship between
ecosystem structure and function and propose a method to examine
potential overlap in remote sensing EBVs. Second, we propose a method
for the selection of remote sensing attributes for developing and
applying EBVs within a single EBV using the independent and shared
relationships of lidar-derived forest structural attributes and modelled
forest structure on ecosystem function. Third, we examine the global
applicability of these results using the relationship between simple and
modelled forest structural attributes with DHIs across a range of
environment gradients and forest types.

@@

British Columbia (BC) is the westernmost province of Canada, and is home
to a variety of terrestrial ecosystems. Located between the Pacific
Ocean and the Rocky Mountains, approximately 64% of the province is
forested, with large environmental and topographic gradients
[@pojar1987; @bcministryofforests2003]. The majority of the forests of
BC are coniferous forests (87.4%), with the remaining forests being
broadleaf (8.9%), wetland-treed (2.7%) and mixed-wood (0.9%)
[@hermosilla2022]. There is a large topographic gradient in the
province, with elevations ranging from sea level to over 4000 m, with
much of the province being mountainous (elevations above 1000 m)
[@valentine1978soil]. Wildfires are a regular disturbance in the
province, with the historical fire regime being disrupted due to
management actions, leading to larger, higher intensity fires becoming
more common in recent years [@brookes2021]. In addition, forestry is a
large industry in the province, with approximately 136000 ha of the
province harvested accounting for \~54 million cubic metres of volume of
wood in 2021 [@canadiancouncilofforestministers2023].

BC is stratified into 16 zones based on the dominant tree species and
climate by the Biogeoclimatic Ecosystem Classification (BEC) system
(@fig-study-map). To examine trends across the large environmental
gradients, we group the BEC zones into five broad biomes: the Nouthern
interior, Sorthern interior, Montane, Alpine, and Coastal groups,
similar to @hamann2006. We also report the average climate data for BEC
zones from 1991-2020, according to @wang2016 (Table 1).

@@

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Precipitation (mm/yr)",
    tmax = "Max Temperature (°C)",
    tmin = "Min Temperature (°C)"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

### National Terrestrial Ecosystem Monitoring System (NTEMS)

#### Best-Available-Pixel

The forest structure, forest disturbance, and land cover layers are
based on a 30-m best-available-pixel (BAP) composite generated by
@hermosilla2016 using the Composite2Change approach. The composites were
developed for each year from 1984-2019 by selecting the best available
imagery from the available growing season Landsat observations, removing
pixels with clouds, cloud shadows, and haze. The methodology uses the
scoring method from @white2014 to select the best pixel for a given year
from Landsat-5 Thematic Mapper, Landsat-7 Enhanced Thematic Mapper Plus,
and Landsat-8 Operational Land Imager imagery. To reduce noise in the
BAP between years, a spectral trend analysis on the Normalized Burn
Ratio was conducted on each pixel, which removes unscreened noise,
detects changes, and temporally interpolates data gaps. This resulted in
a gap-free surface-reflectance composite across the entirety of Canada
[@hermosilla2015]. We utilize data from 2015 for all NTEMS data, except
forest disturbances, where we use all available years of data in the
terrestrial area of BC, Canada.

@@

We used a forest disturbance layer generated by @hermosilla2015a to
restrict the analysis to stands older than 35 years as productivity and
structure metrics have been shown to be strongly decoupled in early
regenerative stands [@bolton2017], and also excluding recently
anthropogenically disturbed stands. Changes were detected using spectral
trend analysis and were attributed using an object-based analysis
approach to a disturbance agent (i.e., wildfire, harvest) using a Random
Forests classification model [@hermosilla2015a]. Changes were detected
with an overall accuracy of 89%, attributed with an accuracy of 92%, and
assigned to the correct year with an accuracy of 89% [@hermosilla2016].

@@

We used a land cover map generated by @hermosilla2022 to stratify our
samples alongside the BEC zones. The 30-m spatial resolution land cover
map for Canada's forested ecosystems was generated using the Virtual
Land Cover Engine framework [@hermosilla2022]. This framework uses a
Random Forests based classification approach, and integrates logical
land cover transitions using a Hidden Markov Model to generate 12 land
cover classes. Training samples were locally selected using an inverse
distance weighting approach [@hermosilla2022]. The treed land cover
classes are broadleaf, coniferous, mixed wood, and wetland-treed, which
we use as a stratification. The land cover classification reached an
overall accuracy of 77.9% ± 4% [@hermosilla2022].

@@

This research was funded by NSERC support of Coops (RGPIN-2018-03851).
Remote sensing data products utilized in this research are free and open
and available for download at <https://ca.nfis.org/maps_eng.html>. We
thank Dr. Michael Wulder and Dr. Joanne White for development and early
access to these National Terrestrial Ecosystem Mapping System (NTEMS)
products. We thank Dr. Elena Razenkova for early access to the
Landsat-derived Dynamic Habitat Indices.

>>>>

---
title: Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.
author:
  - name: Evan R. Muise
    email: evan.muise@student.ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
    attributes:
        corresponding: true
  - name: Margaret E. Andrew
    email: M.Andrew@murdoch.edu.au
    affiliations: 
        - id: murd
          name: Murdoch University
          department: Environmental and Conservation Sciences and Harry Butler Institute
          address: 90 South St
          city: Murdoch, WA, Australia
          postal-code: 6150
  - name: Nicholas C. Coops
    email: nicholas.coops@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
  - name: Txomin Hermosilla
    email: txomin.hermosillagomez@NRCan-RNCan.gc.ca
    affiliations: 
        - id: cfs
          name: Natural Resources Canada
          department: Canada Forest Service (Pacific Forestry Centre)
          address: 506 Burnside Rd W
          city: Victoria, BC, Canada
          postal-code: V8Z 1M5
  - name: A. Cole Burton
    email: cole.burton@ubc.ca
    affiliations: 
        - id: ubc
          name: University of British Columbia
          department: Forest Resources Management
          address: 2424 Main Mall
          city: Vancouver, BC, Canada
          postal-code: V6T 1Z4
  - name: Stephen S. Ban
    email: Stephen.Ban@gov.bc.ca
    affiliations: 
        - id: bcp
          name: Ministry of Environment and Climate Change Strategy
          department: BC Parks
          address: 525 Superior Street
          city: Victoria, BC, Canada
          postal-code: V8V 1T7
keywords: 
  - remote sensing
  - landsat
  - forest structure
  - forest productivity
  - dynamic habitat indices (DHIs)
  - essential biodiversity variables
  - redundancy analysis (RDA)
date: last-modified
bibliography: [bibliography.bib, packages.bib]
csl: scientific-reports.csl
format:
  elsevier-pdf:
    keep-tex: true
    keep-md: true
    journal:
      name: Nature Scientific Reports
      formatting: review
      model: 3p
      cite-style: authoryear
      layout: twocolumn
  docx: 
    reference-doc: my-styles.docx
    keep-md: true
  pdf: 
    pdf-engine: xelatex
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
fig-width: 300
number-sections: true
nocite: |
  @matasci2018, @matasci2018b
---

@@

The Essential Biodiversity Variables (EBV) framework has been proposed
as a monitoring system of standardized, comparable variables that
represents a minimum set of biological information to monitor
biodiversity change at large spatial extents. Six classes of EBVs
(genetic composition, species populations, species traits, community
composition, ecosystem structure and ecosystem function) are defined, a
number of which are ideally suited to observation and monitoring by
remote sensing systems. We used moderate-resolution remotely-sensed
indicators representing two ecosystem level EBV classes (ecosystem
structure and function) to assess their complementarity and redundancy
across a range of ecosystems encompassing significant environmental
gradients. Redundancy analyses found that remote sensing indicators of
forest structure were not strongly related to indicators of ecosystem
productivity (represented by the Dynamic Habitat Indices, DHIs), with
the structural information only explaining 16% of the variation in the
DHIs. Complex metrics of forest structure, such as above ground biomass,
did not contribute additional information over simpler height-based
attributes that can be directly estimated with light detection and
ranging (LIDAR) observations. With respect to ecosystem conditions, we
found that forest types and ecosystems dominated by coniferous trees had
less redundancy between the remote sensing indicators when compared to
broadleaf or mixed forest types. Likewise, higher productivity
environments exhibited the least redundancy between indicators, in
contrast to more environmentally stressed regions. We suggest that
biodiversity researchers continue to exploit multiple dimensions of
remote sensing data given the complementary information they provide on
structure and function focused EBVs, which makes them jointly suitable
for monitoring forest ecosystems.

@@

Monitoring the changing state of biodiversity globally is key to
understanding and mitigating the increased extinction risk for many
species [@thomas2004; @urban2015] and the homogenization of biotic
communities at various spatial and temporal scales
[@mcgillFifteenFormsBiodiversity2015]. The Group on Earth Observations
Biodiversity Observation Network has developed the Essential
Biodiversity Variables (EBVs) [@pereira2013], which are standardized,
comparable variables that represents a minimum set of biological
information required for monitoring biodiversity change at large spatial
extents [@skidmore2021]. The EBVs are designed as an analog to the
Essential Climate Variables framework [@bojinski2014] to be global in
scope, relevant to biodiversity information, feasible to implement and
use, and complementary to one another [@skidmore2021]. There are six
classes of EBVs collected at various scales. The ecosystem EBV classes
(ecosystem structure and ecosystem function) are well suited to be
monitored across large areas using remote sensing technologies
[@skidmore2021]. This has led to a proliferation of potential EBV
datasets that have been generated to describe these two ecosystem EBV
classes, especially in forested environments.

Ecosystem structure is the composition, abundance, and spatial
arrangement of different ecosystem components, and is indicative of
habitat quality and connectivity [@noss1990]. Forest ecosystem structure
is defined as the geographic distribution of trees or biomass, with
forest structural diversity encompassing the measurable morphological
components of the forests (e.g., canopy cover, canopy height, and
structural complexity) [@atkins2023; @valbuena2020]. Forest structural
diversity has been linked to species richness at various spatial scales,
ranging from individual plots to landscapes [@bergen2009; @gao2014]. A
variety of forest structural attributes can be extracted directly from
lidar (light detection and ranging) data, ranging in complexity from
simple (e.g., canopy cover, canopy height) to more complex (e.g.,
foliage height diversity, leaf area index), while others can be
indirectly modelled from lidar and auxiliary data (e.g., aboveground
biomass, basal area) [@coops2021]. A suite of these, and other,
structural attributes have been used as indicators of biodiversity at
local to landscape scales [@lefsky1999; @guo2017; @coops2016;
@bergen2009]. Increased forest structural complexity has been
hypothesized to create additional niches, leading to increased species
diversity [@bergen2009; @heidrich2020], which has been frequently
demonstrated using avian species diversity metrics [@macarthur1961] and
lidar [@clawges2008], or a combination of lidar and satellite imagery
[@herniman2020], with some research indicating that canopy vertical
distribution was the strongest predictor of species richness
[@goetz2007]. Although lidar datasets have traditionally been local in
extent, advances in satellite remote sensing processing have allowed the
extension of three-dimensional forest structure attributes nationally in
a wall-to-wall fashion. The computation of these attributes often
requires data fusion approaches involving lidar data and medium spatial
resolution optical imagery [@coops2021; @matasci2018; @matasci2018b].

Ecosystem functions are measures of ecosystem performance that are the
consequence of one or multiple ecosystem processes [@pettorelli2018;
@ecosyste2005]. Within the EBV framework, vegetation phenology --- the
timing and duration of leaf-on conditions --- and ecologically available
energy --- the amount of accessible and useable energy available for
growth in an ecosystem --- are useful predictors of species richness and
abundances at various scales [@chase2002; @radeloff2019; @coops2019;
@razenkova2023]. Ecologically available energy is the amount of energy
produced in an ecosystem, which is sensitive to limitations such as
recent disturbances and the water and nutrient supplies, and is
measurable via productivity [@bonn2004; @nguyen2022]. This differs from
ambient energy in that it goes beyond the climatic conditions of an
area, and concerns the amount of resources available for consumption in
an ecosystem [@hawkins2003]. Clear links have been developed between
remote sensing-derived spectral reflectance, particularly in the
near-infrared and the red parts of the electromagnetic spectrum
[@rouse1974], which relate to pigment concentrations and leaf structure,
respectively, as well as productivity [@yoder1994]. Consequently, the
normalized difference vegetation index (NDVI) and other vegetation
indices have been correlated with productivity measures [@huang2021;
@sims2006]. The relationship however, is typically non-linear,
principally due to saturation in the spectral response at moderate to
high leaf area index values [@myneni1994]. As a result, the strength of
the relationship between spectral indices and productivity depends on
the ecosystem type, land cover type, amount of canopy cover, and
productivity in the region [@shi2017]. Likewise, relationships between
productivity and biodiversity also contain uncertainty, recognizing that
biodiversity is governed by a variety of factors of which productivity
is but one [@skidmore2021]. Despite these uncertainties, remotely sensed
vegetation indices have enabled the assessment of biodiversity patterns
at single time points [@bonn2004; @nagendra2010; @phillips2008], or
through time when using time series data [@berry2007; @radeloff2019;
@coops2019]

While biodiversity is not solely driven by productivity, the
relationship between productivity and biodiversity can be explained
through various hypothesized mechanisms: including available energy,
environmental stress, and environmental stability. These three
hypotheses have been used to guide the development of a set of remote
sensing indicators, the Dynamic Habitat Indices (DHIs), which summarize
intra-annual dynamics in productivity metrics and have been evaluated as
broad-scale biodiversity indicators globally for a variety of taxa
[@berry2007; @coops2008; @radeloff2019]. According to the available
energy hypothesis [@wright1983; @bonn2004] an increase in ecologically
available energy results in higher species richness. In the context of
the DHIs, the total energy available within a specific area is
quantified using the cumulative DHI, which sums each productivity
observation throughout the course of a year. The environmental stress
hypothesis [@currie2004] suggests that higher levels of minimum
available energy contribute to greater species richness. The minimum
DHI, derived from the lowest productivity observed over the course of a
year, indicates the level of available energy during stressful periods.
The environmental stability hypothesis [@williams2008] proposes that
lower energy variation throughout the year leads to increased species
richness due to reduced resource bottlenecks. The variation DHI captures
the coefficient of variation in a vegetation index over the course of a
year, reflecting the prevalence of resource bottlenecks. The DHIs have
previously been produced at continental to global extents using Moderate
Resolution Imaging Spectroradiometer (MODIS) imagery, while recent
research has focused on producing the DHIs at finer spatial resolutions
using multi-annual time series of Landsat imagery
[@razenkovaMediumresolutionDynamicHabitatInPress].

The DHIs have been shown to be correlated with aspects of biodiversity
across a variety of spatial scales and locations, including Moose
occurrence and abundance in Ontario, Canada [@WOS:000343392200016] and
across Russia [@razenkova2020], beta diversity in butterflies in Canada
[@andrew2012], avian diversity in the coterminous U.S.
[@WOS:000659187700008], Ontario [@coops2009bird] and France
[@bonthoux2018], and global alpha diversity [@radeloff2019]. The DHIs
have also been used to construct novel ecoregionalizations [@andrew2013;
@coops2009]. DHIs, however, are not able to capture all components of
biodiversity due to the many environmental and landscape processes not
observable from satellite data [@coops2009bird].

The suite of remote sensing indicators used in biodiversity monitoring
should be carefully considered. Ideally, indicators should be selected
that are closely related to the phenomenon of interest to allow for
direct integration of monitoring results with management actions
[@pressey2021]. With the advent of large-extent monitoring methods like
satellite remote sensing which have led to a proliferation of potential
EBV datasets, it becomes important to assess the interrelationships
between these datasets and the complementary of the information they
provide to reduce redundancy in future EBV development efforts
[@pereira2013; @skidmore2021]. Identifying and exploring the strong
relationships that may exist between indicators of EBVs may help
elucidate the ecological relationships between these facets of
biodiversity, such as ecosystem structure and function, at local to
global scales. On the other hand, when remote sensing-derived EBVs are
not related, they may be well suited to be used in monitoring programs
together, as complementary sets of observations.

The relationship between ecosystem structure and function have been
examined within a remote sensing context for decades [@huete2002;
@knyazikhin1998; @myneni1994]. Hypothesized mechanisms such as niche
complementarity suggest that productivity increases with stand
structural complexity [@zhang2012], while asymmetric competition for
light can reduce forest productivity with increased structural
complexity [@bourdier2016]. Some aspects of forest structure, such as
canopy cover and leaf area index, also directly influence vegetation
indices, which are used to estimate productivity [@pettorelli2005].
Often, these large-extent estimates of structure and function metrics
are generated from similar data sources, such as the Landsat series of
satellites [@razenkovaMediumresolutionDynamicHabitatInPress ;
@radeloff2019; @matasci2018; @matasci2018b]. Due to this shared data
provenance, and the linkages between biodiversity and both structure and
function, it is important to assess how much unique information these
structure and function metrics represent when used as indicators of
biodiversity.

The goal of this study is to assess the complementarity of two remote
sensing datasets often proposed as indicators of ecosystem structure and
function EBVs. We interpret weak or absent relationships between metrics
to constitute complementary information. To do so, we synthesize
wall-to-wall forest structure data previously generated by modelling
lidar-derived forest structure metrics with medium spatial resolution
satellite imagery using a k-NN imputation method [@matasci2018;
@matasci2018b], alongside a well-established remote sensing-derived
index of ecosystem function (DHIs) [@coops2008; @berry2007;
@radeloff2019; @razenkovaMediumresolutionDynamicHabitatInPress]. First,
we analyze remote sensing indicator complementarity by examining the
multivariate relationship between ecosystem structure and function.
Second, we propose a method for selecting which remote sensing
attributes to develop and apply within a single EBV class by determining
the independent and shared contributions of lidar-derived forest
structural attributes and modelled forest structure to explaining
ecosystem function. Third, we examine the global applicability of these
results by evaluating the consistency of the observed forest
structure-function relationships across a broad range of environment
gradients and forest types.

@@

British Columbia (BC) is the westernmost province of Canada, and is home
to a variety of terrestrial ecosystems. Located between the Pacific
Ocean and the Rocky Mountains, approximately 64% of the province is
forested, with large environmental and topographic gradients
[@pojar1987; @bcministryofforests2003]. The majority of the forests of
BC are coniferous forests (87.4%), with the remaining forests being
broadleaf (8.9%), wetland-treed (2.7%) and mixed-wood (0.9%)
[@hermosilla2022]. There is a large topographic gradient in the
province, with elevations ranging from sea level to over 4000 m, with
much of the province being mountainous (elevations above 1000 m)
[@valentine1978soil]. Wildfires are a regular disturbance in the
province, although larger, higher intensity fires have become more
common in recent years due to disruption of the historical fire regime
[@brookes2021]. In addition, forestry is a large industry in the
province, with approximately 136,000 ha of the province harvested
accounting for \~54 million cubic metres of volume of wood in 2021
[@canadiancouncilofforestministers2023].

BC is stratified into 16 zones based on the dominant tree species and
climate by the Biogeoclimatic Ecosystem Classification (BEC) system
(@fig-study-map). To examine trends across the large environmental
gradients, we group the BEC zones into five broad biomes: the Northern
interior, Southern interior, Montane, Alpine, and Coastal groups,
similar to Hamann et al. [@hamann2006]. We also report the average
climate data for BEC zones from 1991-2020, according to Wang et al.
[@wang2016] (Table 1).

@@

bec_tab %>%
  gt() %>%
  cols_label(
    group = "BEC Group",
    zone = "BEC Zone",
    zone_nm = "Full Name",
    ppt = "Average Monthly Precipitation (mm/yr)",
    tmax = "Average Monthly Max Temperature (°C)",
    tmin = "Average Monthly Min Temperature (°C)"
  ) %>%
  tab_header(title = "Biogeoclimatic Ecosystem Classification (BEC) Zones, their aggregated biomes, and their monthly average climate values for precipitation, maximum temperature, and minimum temperature. Climate data from Wang et al. 2016 averaged for 1991-2020.")
```

@@

### National Terrestrial Ecosystem Monitoring System data products

#### Surface-reflectance image composites

The forest structure, forest disturbance, and land cover layers are
based on a 30-m best-available-pixel (BAP) composite generated by
Hermosilla et al. @hermosilla2016 using the Composite2Change approach.
The composites were developed for each year from 1984-2019 by selecting
the best available imagery from the available growing season Landsat
observations, removing pixels with clouds, cloud shadows, and haze. The
methodology uses the scoring method from White et al. @white2014 to
select the best pixel for a given year from Landsat-5 Thematic Mapper,
Landsat-7 Enhanced Thematic Mapper Plus, and Landsat-8 Operational Land
Imager imagery. A spectral trend analysis was conducted over these
initial BAP composites using the Normalized Burn Ratio (a vegetation
index) on each pixel, to remove unscreened noise, detect changes, and
temporally interpolate data gaps. This resulted in a gap-free
surface-reflectance composite across Canada [@hermosilla2015]. We use
data from 2015 for all National Terrestrial Ecosystem Monitoring System
products, except forest disturbances, which comprise the period
1985-2020.

@@

We used the forest disturbance layer generated by Hermosilla et al.
@hermosilla2016 to restrict the analysis to stands older than 35 years
as productivity and structure metrics have been shown to be strongly
decoupled in early regenerative stands [@bolton2017], and to exclude
recently anthropogenically disturbed stands. The disturbances detected
using spectral trend analysis were attributed to a disturbance agent
(i.e., wildfire, harvest) within an object-based analysis approach using
a Random Forests classification model [@hermosilla2015a]. Changes were
detected with an overall accuracy of 89%, attributed with an accuracy of
92%, and assigned to the correct year with an accuracy of 89%
[@hermosilla2016].

@@

We used the land cover map for 2015 generated by Hermosilla et al.
@hermosilla2022 to stratify our samples alongside the BEC zones. The
30-m spatial resolution land cover map for Canada's forested ecosystems
was generated using the Virtual Land Cover Engine framework
[@hermosilla2018]. This framework uses a Random Forests based
classification approach, and integrates logical land cover transitions
using a Hidden Markov Model to generate 12 land cover classes. The land
cover classification reached an overall accuracy of 77.9% ± 4%
[@hermosilla2022]. We restricted our sample to the four treed land cover
classes: broadleaf, coniferous, mixed wood, and wetland-treed.

@@

This research was funded by NSERC support of Coops (RGPIN-2018-03851).
Remote sensing data products used in this research are free and open and
available for download at <https://ca.nfis.org/maps_eng.html>. We thank
Dr. Michael A. Wulder and Dr. Joanne C. White for development and early
access to these National Terrestrial Ecosystem Mapping System (NTEMS)
products. We thank Dr. Elena Razenkova for early access to the
Landsat-derived Dynamic Habitat Indices.

# Author Contributions

Conceptualization: ERM, MEA, NCC, ACB and SSB. Methodology: ERM, MEA,
NCC and TH. Software: ERM and MEA. Formal Analysis: ERM, MEA, and NCC.
Investigation: ERM. Writing - Original Draft: ERM. Writing - Review and
Editing: ERM, MEA, NCC, TH, ACB and SSB. Visualization: ERM, MEA, TH,
NCC. Supervision: NCC. All authors approve the submitted version of the
manuscript.

# Competing Interests

MEA is an Editorial Board Member of *Scientific Reports*. The authors
declare no other competing interests.

# Data Availability
